\pkg{depmixS4} : An \proglang{R}-package for hidden Markov models

### 2010-03-04 papers/jss/dpx4Rev.Rnw
- identity link now default in init models - minor changes to the jss article
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@377 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

For the example data above, $b_j^k$ could be a Gaussian distribution
function for the response time variable, and a Bernoulli distribution
for the accuracy variable.  In the models we are considering here,
both the transition probability functions $a_{ij}$ and the initial
state probability functions $\greekv{\pi}$ may depend on covariates as
well as the response distributions $b_{j}^{k}$.

@@

Response times are a very common dependent variable in psychological
experiments and hence form the basis for infernce about many
psychological processes.  A potential threat to such inference based
on response times is formed by the speed-accuracy trade-off: different
participants in an experiment may respond differently to typical
instructions to `respond as fast and accurate as possible'.  A popular
model which takes the speed-accuracy trade-off into account is the
diffusion model \citep{Ratcliff1978}, which has proven to provide
accurate descriptions of response times in many different settings.

One potential problem with the diffusion model is that it predicts a
continuous trade-off between speed and accuracy of responding, i.e.,
when participants are pressed to respond faster and faster, the
diffusion model predicts that this would lead to a gradual decrease in
accuracy.  The \code{speed} data set that we analyze below was
gathered to test this hypothesis versus the alternative hypothesis
stating that there is a sudden transition from slow and accurate
responding to fast responding at chance level.  At each trial of the
experiment, the participant is shown the current setting of the
relative reward for speed versus accuracy.  The bottom panel of
figure~ref{fig:speed} shows the values of this variable.  The main
question in this experiment was what would happen when this reward
variable would change from reward for accuracy only to reward for
speed only.  The \code{speed} data that we analyse here are from
participant A in Experiment 1 in \citet{Dutilh2010}, who provide a
complete description of the experiment and the relevant theoretical
background.

The central hypothesis about this data is whether indeed it is best
described by two modes of responding rather than a single mode of
responding with a continuous trade-off between speed and accuracy.
The hallmark of having a discontinuity between slow versus speeded
responding is that the switching between the two modes is assymetric
\citep[see e.g.][for a theoretical underpinning of this
claim]{Maas1992}.  The \code{fit} help page of \pkg{depmixS4} provides
a number of examples in which the assymetricity of the switching
process is tested; those examples and other candidate models are
discussed at length in \citet{Visser2009b}.

@@

\paragraph{Starting values} Note also that start values for the
transition parameters are provided in this call using the
\code{trstart} argument.  At this time, the package does not provide
automatic starting values.  Starting values for parameters can be
provided using three arguments: \code{instart} for the parameters
relating to the prior probabilities, \code{trstart}, relating the
transition models, and \code{respstart} for the parameters of the
response models.  Note that the starting values for the initial and
transition models as well as of the multinomial logit response models
are interpreted as {\em probabilities}, and internally converted to
multinomial logit parameters.

@@

These statistics can also be extracted using \code{logLik}, \code{AIC}
and \code{BIC}, respectively.  By comparison, a 1-state model for
these data, i.e. assuming there is no mixture, has a log-likelihood of
-305.33, and 614.66, and 622.83 for the AIC and BIC respectively.
Hence, the 2-state model fits the data much better than the 1-state
model.  Note that the 1-state model can be specified using \code{mod <-
depmix(rt~1, data=speed, nstates=1)}, although this model is trivial
as it will simply return the mean and the sd of the rt variable.

@@

By default, the transition probabilities and the initial state
probabilities are parameterized using the multinomial logistic model.
More precisely, each row of the transition matrix is parameterized by
a baseline category logistic multinomial, meaning that the parameter
for the base category is fixed at zero \citep[see][p.\ 267 ff., for
multinomial logistic models and various
parameterizations]{Agresti2002}.  The default baseline category is the
first state.  Hence, for example, for a 3-state model, the initial
state probability model would have three parameters of which the first
is fixed at zero and the other two are freely estimated.

The multinomial logistic model allows us to include covariates on the
initial state and transition probabilities.  \citet{Chung2007} discuss
a related latent transition model for repeated measurement data
($T=2$) using logistic regression on the transition parameters; they
rely on Bayesian methods of estimation.  Covariates on the transition
probabilities can be specified using a one-sided formula as in the
following example:
<<>>=
set.seed(1)
mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(), 
    transition=~scale(Pacc), instart=runif(2))
fm <- fit(mod,verbose=FALSE)
@

@@

Multivariate data can be modelled by providing a list of formulae as
well as a list of family objects for the distributions of the various
responses.  In above examples we have only used the response times
which were modelled with the Gaussian distribution.  The accuracy
variable in the \code{speed} data can be modelled with a multinomial
by specifying the following:
<<>>=
set.seed(1)
mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
    family=list(gaussian(),multinomial()),
    transition=~scale(Pacc),instart=runif(2))
fm <- fit(mod,verbose=FALSE)
@

@@

The balance scale taks is a famous task for testing cognitive
strategies developed by Jean Piaget \citep[see]{Siegler1981}.
Figure~\ref{fig:balance} provides an example of a balance scale item.
Participants' task is to say to which side the balance will tip when
released, or alternatively, whether it will stay in balance.  The item
shown in Figure~\ref{fig:balance} is a so-callled distance item: the
number of weights placed on each side is equal, and only the distance
of the weights to the fulcrum differs between each side.

@@

Note here that we define a \code{mix} model instead of a \code{depmix}
models as these data form independent observations.  More formally,
\code{depmix} models extend the class of \code{mix} models by adding
the transition models.  As for fitting \code{mix} models: as can be
seen in equation~\ref{eq:Q}, the EM algorithm can be applied by simply
dropping the second summand containing the transition parameters, and 
this is implemented as such in the EM algorithms in \pkg{depmixS4}.

@@

Using package \pkg{Rdonlp2} by \citet{Tamura2009} or \pkg{Rsolnp}
\citet{Ghalanos2010}, parameters may be fitted subject to general linear
(in-)equality constraints.  Constraining and fixing parameters is done
using the \code{conpat} argument to the \code{fit}-function, which
specifies for each parameter in the model whether it's fixed (0) or
free (1 or higher).  Equality constraints can be imposed by having two
parameters have the same number in the \code{conpat} vector.  When
only fixed values are required, the \code{fixed} argument can be used
instead of \code{conpat}, with zeroes for fixed parameters and other
values (ones e.g.) for non-fixed parameters.  Fitting the models
subject to these constraints is handled by the optimization routine
\code{donlp2}.  To be able to construct the \code{conpat} and/or
\code{fixed} vectors one needs the correct ordering of parameters
which is briefly discussed next before proceeding with an example.

\paragraph{Parameter numbering} When using the \code{conpat} and
\code{fixed} arguments, complete parameter vectors should be supplied,
i.e., these vectors should have length of the number of parameters of
the model, which can be obtained by calling \code{npar(object)}.  Note
that this is not the same as the degrees of freedom used e.g.\ in the
\code{logLik} function because \code{npar} also counts the baseline
category zeroes from the multinomial logistic models.  Parameters are
numbered in the following order:
\begin{enumerate}
	\item  the prior model parameters
	\item  the parameters for the transition models
	\item  the response model parameters per state (and subsequently
	per response in the case of multivariate time series)
\end{enumerate}

@@

User-defined distributions should extend the \code{response}-class and
have the following slots:
\begin{enumerate}
	\item y: the response variable
	\item x: the design matrix, possibly only an intercept
	\item paramaters: a named list with the coefficients and possibly 
	other parameters, e.g., the standard deviation in the Gaussian 
	response model
	\item fixed: a vector of logicals indicating whether parameters are 
	fixed
	\item npar: numerical indicating the number of parameters of the model
\end{enumerate}

@@

After defining all the necessary methods for the new response model, 
we can  now define the dependent mixture model using this reponse model. 
The function \code{makeDepmix} is included in \pkg{depmixS4} to have 
full control over model specification, and we need it here.

@@

Using \code{summary} will print the fitted parameters.  Note that the
use of \code{makeDepmix} allows the possibility of, say, fitting a
Gaussian in one state and an exgaus distribution in another state.
Note also that according to the AIC and BIC, the model with the exgaus
describes the data much better than the same model in which the
response times are modeled as gaussian.

@@

\pkg{depmixS4} provides a flexible framework for fitting dependent
mixture models for a large variety of response distributions.  It can
also fit latent class regression and finite mixture models, although
it should be noted that more specialized packages are available for
this such as \pkg{FlexMix} \citep{Leisch2004}.  The package is
intended for modeling of (individual) time series data with the aim of
characterizing the transition processes underlying the data.  The
possibility to use covariates on the transition matrix greatly
enhances the flexibility of the model.  The EM algorithm uses a very
general interface that allows easy addition of new response models.

@@

Ingmar Visser was supported by an EC Framework 6 grant, project 516542
(NEST).  Maarten Speekenbrink was supported by the ESRC Centre for
Economic Learning and Social Evolution (ELSE).  Han van der Maas
provided the speed-accuracy data \cite{Dutilh2010} and thereby
neccessitated implementing models with time-dependent covariates.
Brenda Jansen provided the balance scale data set \citep{Jansen2002}
which was the perfect opportunity to test the covariates on the prior
model parameters.  The examples in the help files use both of these
data sets.

>>>>

For the example data above, $b_j^k$ could be a Gaussian distribution
function for the response time variable, and a Bernoulli distribution
for the accuracy variable.  In the models considered here,
both the transition probability functions $a_{ij}$ and the initial
state probability functions $\greekv{\pi}$ may depend on covariates as
well as the response distributions $b_{j}^{k}$.

@@

Response times are a very common dependent variable in psychological
experiments and hence form the basis for inference about many
psychological processes.  A potential threat to such inference based
on response times is formed by the speed-accuracy trade-off: different
participants in an experiment may respond differently to typical
instructions to `respond as fast and accurate as possible'.  A popular
model which takes the speed-accuracy trade-off into account is the
diffusion model \citep{Ratcliff1978}, which has proven to provide
accurate descriptions of response times in many different settings.

One potential problem with the diffusion model is that it predicts a
continuous trade-off between speed and accuracy of responding, i.e.,
when participants are pressed to respond faster and faster, the
diffusion model predicts that this would lead to a gradual decrease in
accuracy.  The \code{speed} data set that we analyze below was
gathered to test this hypothesis versus the alternative hypothesis
stating that there is a sudden transition from slow and accurate
responding to fast responding at chance level.  At each trial of the
experiment, the participant is shown the current setting of the
relative reward for speed versus accuracy.  The bottom panel of
figure~\ref{fig:speed} shows the values of this variable.  The 
experiment was designed to investigate what would happen when this reward
variable changes from reward for accuracy only to reward for
speed only.  The \code{speed} data that we analyse here are from
participant A in Experiment 1 in \citet{Dutilh2010}, who provide a
complete description of the experiment and the relevant theoretical
background.

The central question regarding this data is whether it is indeed best
described by two modes of responding rather than a single mode of
responding with a continuous trade-off between speed and accuracy.
The hallmark of a discontinuity between slow versus speeded
responding is that switching between the two modes is asymmetric
\citep[see e.g.][for a theoretical underpinning of this
claim]{Maas1992}.  The \code{fit} help page of \pkg{depmixS4} provides
a number of examples in which the asymmetry of the switching
process is tested; those examples and other candidate models are
discussed at length in \citet{Visser2009b}.

@@

\paragraph{Starting values} Note also that start values for the
transition parameters are provided in this call by the
\code{trstart} argument.  At this time, the package does not provide
automatic starting values.  Starting values for parameters can be
provided using three arguments: \code{instart} for the parameters
relating to the prior probabilities, \code{trstart}, relating the
transition models, and \code{respstart} for the parameters of the
response models.  Note that the starting values for the initial and
transition models as well as multinomial logit response models
are interpreted as {\em probabilities}, and internally converted to
multinomial logit parameters (if a logit link function is used).

@@

These statistics can also be extracted using \code{logLik}, \code{AIC}
and \code{BIC}, respectively.  By comparison, a 1-state model for
these data, i.e. assuming there is no mixture, has a log-likelihood of
$-305.33$, and 614.66, and 622.83 for the AIC and BIC respectively.
Hence, the 2-state model fits the data much better than the 1-state
model.  Note that the 1-state model can be specified using \code{mod <-
depmix(rt~1, data=speed, nstates=1)}, although this model is trivial
as it will simply return the mean and the sd of the rt variable.

@@

By default, the transition probabilities and the initial state
probabilities are parameterized using a multinomial model with an 
identity link function.
Using a multinomial logistic model allows us to include covariates on the
initial state and transition probabilities. In this case, each row of the 
transition matrix is parameterized by
a baseline category logistic multinomial, meaning that the parameter
for the base category is fixed at zero \citep[see][p.\ 267 ff., for
multinomial logistic models and various
parameterizations]{Agresti2002}.  The default baseline category is the
first state.  Hence, for example, for a 3-state model, the initial
state probability model would have three parameters of which the first
is fixed at zero and the other two are freely estimated. \citet{Chung2007} discuss
a related latent transition model for repeated measurement data
($T=2$) using logistic regression on the transition parameters; they
rely on Bayesian methods of estimation.  Covariates on the transition
probabilities can be specified using a one-sided formula as in the
following example:
<<>>=
set.seed(1)
mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(), 
    transition=~scale(Pacc), instart=runif(2))
fm <- fit(mod,verbose=FALSE)
@

@@

Multivariate data can be modelled by providing a list of formulae as
well as a list of family objects for the distributions of the various
responses.  In above examples we have only used the response times
which were modelled as a Gaussian distribution.  The accuracy
variable in the \code{speed} data can be modelled with a multinomial
by specifying the following:
<<>>=
set.seed(1)
mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
    family=list(gaussian(),multinomial()),
    transition=~scale(Pacc),instart=runif(2))
fm <- fit(mod,verbose=FALSE)
@

@@

The balance scale taks is a famous task for testing cognitive
strategies developed by Jean Piaget \citep[see][]{Siegler1981}.
Figure~\ref{fig:balance} provides an example of a balance scale item.
Participants' task is to say to which side the balance will tip when
released, or alternatively, whether it will stay in balance.  The item
shown in Figure~\ref{fig:balance} is a so-callled distance item: the
number of weights placed on each side is equal, and only the distance
of the weights to the fulcrum differs between each side.

@@

Note here that we define a \code{mix} model instead of a \code{depmix}
model as these data form independent observations.  More formally,
\code{depmix} models extend the class of \code{mix} models by adding
transition models.  As for fitting \code{mix} models: as can be
seen in Equation~\ref{eq:Q}, the EM algorithm can be applied by simply
dropping the second summand containing the transition parameters, and 
this is implemented as such in the EM algorithms in \pkg{depmixS4}.

@@

Using package \pkg{Rdonlp2} by \citet{Tamura2009} or \pkg{Rsolnp}
\citet{Ghalanos2010}, parameters may be fitted subject to general linear
(in-)equality constraints.  Constraining and fixing parameters is done
using the \code{conpat} argument to the \code{fit}-function, which
specifies for each parameter in the model whether it's fixed (0) or
free (1 or higher).  Equality constraints can be imposed by giving two
parameters the same number in the \code{conpat} vector.  When
only fixed values are required, the \code{fixed} argument can be used
instead of \code{conpat}, with zeroes for fixed parameters and other
values (ones e.g.) for non-fixed parameters.  Fitting the models
subject to these constraints is handled by the optimization routine
\code{donlp2}.  To be able to construct the \code{conpat} and/or
\code{fixed} vectors one needs the correct ordering of parameters
which is briefly discussed next before proceeding with an example.

\paragraph{Parameter numbering} When using the \code{conpat} and
\code{fixed} arguments, complete parameter vectors should be supplied,
i.e., these vectors should have length equal to the number of parameters of
the model, which can be obtained by calling \code{npar(object)}.  Note
that this is not the same as the degrees of freedom used e.g.\ in the
\code{logLik} function because \code{npar} also counts the baseline
category zeroes from the multinomial logistic models.  Parameters are
numbered in the following order:
\begin{enumerate}
	\item  the prior model parameters
	\item  the parameters for the transition models
	\item  the response model parameters per state (and subsequently
	per response in the case of multivariate time series)
\end{enumerate}

@@

User-defined distributions should extend the \code{response}-class and
have the following slots:
\begin{enumerate}
	\item y: the response variable
	\item x: the design matrix, possibly only an intercept
	\item paramaters: a named list with the coefficients and possibly 
	other parameters (e.g., the standard deviation in the Gaussian 
	response model)
	\item fixed: a vector of logicals indicating whether parameters are 
	fixed
	\item npar: numerical indicating the number of parameters of the model
\end{enumerate}

@@

After defining all the necessary methods for the new response model, 
we can  now define the dependent mixture model using this response model. 
The function \code{makeDepmix} is included in \pkg{depmixS4} to have 
full control over model specification, and we need it here.

@@

Using \code{summary} will print the fitted parameters.  Note that the
use of \code{makeDepmix} allows the possibility of, say, fitting a
Gaussian in one state and an exgaus distribution in another state.
Note also that according to the AIC and BIC, the model with the exgaus
describes the data much better than the same model in which the
response times are modelled as Gaussian.

@@

\pkg{depmixS4} provides a flexible framework for fitting dependent
mixture models for a large variety of response distributions.  It can
also fit latent class regression and finite mixture models, although
it should be noted that more specialized packages are available for
this such as \pkg{FlexMix} \citep{Leisch2004}.  The package is
intended for modelling of (individual) time series data with the aim of
characterizing the transition processes underlying the data.  The
possibility to use covariates on the transition matrix greatly
enhances the flexibility of the model.  The EM algorithm uses a very
general interface that allows easy addition of new response models.

@@

Ingmar Visser was supported by an EC Framework 6 grant, project 516542
(NEST).  Maarten Speekenbrink was supported by ESRC grant RES-062-23-1511 
and the ESRC Centre for Economic Learning and Social Evolution (ELSE).  
Han van der Maas provided the speed-accuracy data \cite{Dutilh2010} and thereby
neccessitated implementing models with time-dependent covariates.
Brenda Jansen provided the balance scale data set \citep{Jansen2002}
which was the perfect opportunity to test the covariates on the prior
model parameters.  The examples in the help files use both of these
data sets.
